Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2024-02-08 paper/markdown/review_response.Rmd

---
title: "Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center"
subtitle: "Point-to-point response"
author: "Psychiatry Study Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: ptsd_biblio.bib
csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 9)

```

\newpage

# Editor

Dear Editor,

Thank you very much for the positive and very thoughtful review of our manuscript. We appreciate a lot the reviewers’ comments and suggestions to improve our manuscript. 
We incorporated them in the revised text.
Please find below the point-to-point response to the reviewers’ issues. 
Since the reviewers requested more information and literature evidence, we also deleted few less important references from the initial submission to keep the bibliography within reasonable limits. 
Additionally, some minor copyediting was done. 
Changes in the main manuscript text are highlighted in red. 
We hope that the manuscript is now suitable for publication in _European Archives of Psychiatry and Clinical Neuroscience_.

Best regards,

Katharina Hüfner

\newpage

# Reviewer 1

## Point 1

Summary: the study provides the first data on mental healthy consequences of alpine sport accidents. 
The topic is of high interest specifically of the medical specialists of the alpine region countries. 
The study design is sound and the methods a highly innovative. 
Overall I recommend publishing this interesting data.

## Response 1

Thank you for the constructive feedback and the overall positive assessment of our study.

## Point 2

__Abstract__: The abstract is well structured. 
Probably it might be better not to make the first sentence about a topic which the study does not cover (positive effects of mountain sports).

## Response 2

We agree that the first sentence of __Abstract__ does not specifically relate to the subject of our research, which concerns primarily negative psycho-pathological burden of mountain sport accidents.
We changed this sentence of __Abstract__ attempting to accommodate the reviewer´s concern but did not completely remove it because we feel that especially “mountain sports” are generally believed to bolster mental health. 
Additionally, individuals performing mountain sports are often portrayed as resilient personalities protected from mental disorders, which we can in part support with our data (__Supplementary Figures S4__ and __S5__). 
In our clinical practice, these beliefs are a common source of stigma in members of the mountaineering community. 
So I hope this makes it more clear why we did not remove the first sentence completely and hopefully the reviewer can understand our intentions.

## Point 3

__Introduction__: The introduction is well written. 
At the end I think it would help to make the hypothesis more clear. 
I know that with SVM methods a hypothesis free approach is common, however the authors intend to (1) categorise the sample (2) find predictors.

## Response 3

Thank you for this suggestion. 
In the last paragraph of __Introduction__, we stressed the exploratory character of the study and the general strategy, i.e. classification of participants by their mental health measures followed by a search for predictors of mental health classes/clusters.

## Point 4

__Methods__: The methods are highly interesting and - as far as I can judge - are sound. 
Sample: The authors clearly state the problem of the low response rate of about 6%. 
There is a clear bias, however studies indicate that in mental health normally prevalences are under reported not overestimated.

## Response 4

We absolutely agree with your comment that the low response rate is a source of bias. 
Unfortunately this is a challenge of many studies tackling with mental disorders such as PTSD, anxiety and depression.

To assess the potential effects of the low response rate on the cohort composition we included an extensive comparison of the analysis cohort with invited individuals and participants excluded due to incomplete psychometric data in the initial submission. 
In the revised manuscript we investigated the potential selection bias in more detail, by comparing sociodemographic features of the cohort with the general Austrian population and by comparing characteristic of the mountain sport accident in the study cohort with a nation-wide collective of Austrian mountain accident victims.

As presented in __Supplementary Table S6__ and discussed in __Results/Characteristic of the study cohort__, the study sample was characterized by significantly higher frequencies of males, persons with tertiary education grade and actively employed as compared with the general Austrian population. 
In turn, rates of smokers and individuals suffering from chronic physical or mental disorders was significantly lower. 
Effect sizes of those differences were, however, very small (Cramer's V 0.00078 to 0.0025). 
Results of the comparison of the accident characteristic are presented in __Supplementary Figure 1__ and described in __Results/Characteristic of the study cohort__. 
They indicate significantly higher frequency of winter sport accidents in the study cohorts than in the Austrian data set. 
Yet the effect size of the difference was again small (Cramer’s V ≤ 0.14). 
Taken together, these analyses suggest that there is indeed a weak but significant selection bias towards men, well educated, professionally active and healthy individuals, and victims of winter sport accidents. 
We point this out in the limitation section of __Discussion__ of the revised manuscript.

To address the possible under- or overestimation (see: __Point 10__, __Reviewer 2__) of mental disorder symptoms in the study cohort, we compared frequency of traumatic events, manifest PTSD, symptoms of anxiety and depression in the study cohort with large survey studies [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017], as well as data of the German mental health monitoring [@Junker2023] and an Austrian general health microcensus study [@StatistikAustria2019a]. 
Frequency of traumatic events and manifest PTSD in our sample was lower or comparable with the estimated prevalence in the general German population reported by Hauffa and colleagues [@Hauffa2011] (__Supplementary Figure S4A__). 
Interestingly, frequency of manifest PTSD and frequency of the PCL-5 domain B, C, D, and E PTSD symptoms did not differ significantly from figures published by Mikutta et al. for Swiss mountain rescuers [@Mikutta2022] (__Supplementary Figure S4B__). 
Furthermore, frequency of 
anxious symptoms (`r lit_plots$depr_anx_stats %>% filter(disorder == 'anxiety', source == 'cohort') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`) assessed by the GAD-2 instrument was significantly higher than in the general German population within the time frame of the study survey 
(`r lit_plots$depr_anx_stats %>% filter(disorder == 'anxiety', source == 'desurv') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`). 
Depressive symptoms measured by the PHQ-2 tool were, in turn, much less prevalent 
in our cohort (`r lit_plots$depr_anx_stats %>% filter(disorder == 'depression', source == 'cohort') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`) than 
estimates for the general German population (`r lit_plots$depr_anx_stats %>% filter(disorder == 'depression', source == 'desurv') %>% mutate(text = paste0(signif(percent, 2), '% [95% CI: ', signif(lower_ci, 2), ' to ', signif(upper_ci, 2), '%]')) %>% .$text`). 
Finally, frequency of moderate-to-severe symptoms of depression (PHQ-8 $\geq$ 10) was comparable in the study samples and the Austrian microcensus results (__Supplementary Figure S5AB__). 
These results let us reason that, despite the mountain accident, the overall mental health status of the study cohort was comparable to the general population of German-speaking countries and we elaborate on this in __Results/Characteristic of the study cohort__ and __Discussion__.

## Point 5

__Discussion__: The discussion is well structured: Probably it would help if you’d point out your own findings at the beginning of each paragraph.

## Response 5

Thank you for the suggestion. 
We summarize our most important findings in the first paragraph of __Discussion__. 
Where appropriate we have now also added additional summaries at the beginning of the Discussion's paragraphs.

## Point 6

Line 51: The think the study was performed at the University of Grenoble which is in France not Switzerland.

## Response 6

We are sorry for the confusion. 
This was corrected in the revised manuscript.

## Point 7

Throughout the discussion: Please find a clear terminology of the PTSD: sometimes you talk about manifest PTSD sometimes about PTSD symptoms. 
As you mostly do refer to symptoms of PTSD I’d stick with that.

## Response 7

Thank you for raising this important issue. 
We agree that the distinction between symptoms of PTSD and manifest PTSD was not completely coherent in the initial text. 
We define symptoms of PTSD as positivity for at least one domain of the PCL-5 tool. 
Manifest PTSD is considered for participants screened positive for all four domains of the PCL-5 instrument [@Bovin2016]. 
This distinction is described now in __Methods__ and __Supplementary Methods__.

Of note, while single PTSD symptoms affected nearly one of five study participants, manifest PTSD was discerned in four individuals only. 
We corrected the ambiguous passages and describe frequency of manifest PTSD solely in __Results/Characteristic of the study cohort__. 
We use the term 'symptoms of PTSD elsewhere in the revised text.

## Point 8

Figures: The figures a very clear and have a nice layout. Congrats.

## Response 8

Thank you very much.

\newpage

# Reviewer 2

## Point 9

This paper aims at characterizing mental health in persons after mountain sport accidents. 
This is an interesting question approached with a very transparent and accessibly documented analysis.

## Response 9

Thank you for the constructive feedback and appreciation of our work!

## Point 10

I would almost certainly expect there to be a strong self-selection bias in the obtained responses to your survey (i.e., those still dealing with the consequences of their accident), likely overestimating the prevalence of mental health-related impairment in these individuals. 
In addition, the supplementary table S4 shows that respondents had more severe injuries requiring hospitalization and surgery compared to non-respondents, also reflecting self-selection of particularly affected individuals among those who experienced a mountaineering accident. Does that change the implication(s) of your results?

## Response 10

We absolutely agree that there is a selection bias in this study. 
This was already pointed out in the limitations section of __Discussion__ of the initial submission. 
As described in more detail in __Response 4__ we took now additional measures to better characterize the potential selection bias. 
From comparisons of our sample with the general Austrian population (__Supplementary Table S4__) and the collective of mountain accident victims in Austria in 2023 (__Supplementary Figure S1__), we infer that the study cohort consisted of significantly more males, highly educated and healthy individuals and victims of winter sport accidents as expected for the general population. 
Effect size of those differences was weak (Cramer's V effect size statistic $\leq$ 0.1)
We elaborate on those findings in __Results/Characteristic of the study cohort__ and summarize these findings on the selection bias in the limitation part of __Discussion__. 
Of note, the selection bias may explain low frequency of traumatic events prior to the mountain accident and rare cases of manifest PTSD in our cohort, as e.g. men and tertiary school graduates are less likely to be exposed to traumatic events and develop PTSD in the lifetime [@Darves-Bornoz2008] (__Supplementary Figure S1__). 
We describe this possibility in __Discussion__ of the revised manuscript.

Next, we addressed the issue of under- or over-reporting of mental health symptoms in the study sample, which may be attributed to the selection bias of the sample. 
To this end, we compared frequency of traumatic events, manifest PTSD, anxious and depressive symptoms in the study cohort with estimates for the general population (__Supplementary Figure S4__). 
As outlined in __Response 4__ in more detail, frequency of traumatic events and manifest PTSD was at the lower tail of the reported prevalence and comparable with figures for the German population published by Hauffa and colleagues [@Hauffa2011]. 
Anxious symptoms measured with the GAD-2 instrument were more common in the study cohort than in the German population during the survey [@Junker2023]. 
In turn, depending on the tool and reference population, overall depressive symptoms and symptoms of moderate-to-severe depression were lower or comparable with the general German or Austrian population (__Supplementary Figure S5__). 
We discuss in __Results/Characteristic of the study cohort__ and in __Discussion__, that the overall frequency of mental disorder symptoms in our cohort is, in spite of the accident, comparable with the general population. 
It is hence possible that bias of the sample towards protective characteristic (education, male sex, economic status) compensates the negative effects of the accident and injury on mental health. 
Additionally, in light of the comparison of the mental disorder symptoms in the cohort and estimates for the general population, over-reporting of mental disorder symptoms in the study cohort is unlikely.

Finally, we investigated the impact of factors related to the selection bias of our sample, i.e. different income structure, more frequent hospitalization, surgery and more severe injury in non-responders or excluded survey participants than in the analysis cohort, on definition of the mental health clusters. 
This analysis was motivated the general concept of one-at-time sensitivity analysis [@Thabane2013]. 
In brief, we developed the mental health clusters with the PAM/cosine distance procedure for the genuine analysis cohort and the following:

* the analysis cohort appended with observations with incomplete psychometry data. The missing values of the mental health assessment battery were imputed with the 9-nearest neighbor algorithm [@Hastie2023].

* the analysis cohort without, respectively, participants with no, low, intermediate and high annual income

* the analysis cohort without participants with, respectively, mild, moderate and severe-to-critical injuries

* separately for hospitalized and non-hospitalized study participants

* referring to your issue of effects of the accident - survey time on mental health status, we developed the clusters for the analysis cohort without, respectively, accidents in 2018, 2019 and 2020

* referring to your issue of multiple mountain sport accidents, we also defined the mental health clusters for participants with and without prior mountain sport accidents

Due to the low number of participants subjected to surgery (n = `r get_percent(ptsd$dataset, 'surgery_done')$count['yes']`), we were not able to define meaningful clusters for this subset. 
Subsequently, we compared fractions of explained clustering variances and cluster sizes between the mental clusters of the genuine analysis cohort and the mental clusters defined for the extended or depleted data sets. 
As shown in __Supplementary Figure S10__, inclusion of the missing observations or removal of subsets of the analysis cohort had minimal effects on the clustering quality measured by the explained variance statistic (analysis cohort: V = `r summary(se_globals$clust_obj)$frac_var %>% signif(2)`, extended/depleted data sets: V = `r filter(se_summary$stats, split_factor %in% c('missing', 'household_income_class', 'prior_accident', 'hospitalization', 'injury_sev_strata'))$frac_var %>% range %>% signif(2) %>% paste(collapse = ' to ')`), or on the cluster size distribution. 
Hence, the effects of exclusion of participants with incomplete psychometric data, changes in the income structure, injury severity, hospitalization status, prior accidents or particular accident years had little effect on clustering of the study cohort by its mental health features.

Interestingly, participants excluded from the analysis due to incomplete psychometric data, were likely characterized by low scoring of post-traumatic growth, low resilience and quality of life and high scores of panic as compared with participants with complete psychometric data and were hence predominantly assigned to the PTS cluster (__Supplementary Figure S11__). 
This observation suggests that by exclusion of survey participants with incomplete mental health data, rates of poor mental health and life quality are rather underestimated in our analysis. 
We elaborate on this new data in __Results/Three clusters of mental health response after mountain sport accidents__ and __Discussion__.

## Point 11

Respondents completed the survey after about 3.5 years (median) after their accident. 
This seems like a long time, how certain are you that any mental health-related burden is still due to the accident? 
Did you screen for any events that might have happened in the meantime?

## Response 11

We agree that the accident - survey interval may be problematic. 
Our initial intention was to analyze persistent, long term alterations of mental health in victims of mountain sport accidents, i.e. at least six months after the accident - we state this point clearly in the final paragraph of __Introduction__ and __Methods/Analysis endpoints__.

During the exploratory data analysis, we have not found any effects of the accident-survey time on the psychometric scores and for effects of the accident year on the cluster assignment (__Review Figure R\@ref(fig:fig-time)__). 
Results of this analysis are available for an interested reader in the [GitHub code repository of the project](https://github.com/PiotrTymoszuk/mental_accident/tree/main/paper). 
Furthermore, we were not able to identify any substantial effects on accidents in 2018, 2019 and 2020 on development of the mental clusters in a sensitivity analysis described in __Response 10__ (__Supplementary Figure S10__). 
In light of those data, we are pretty confident that the time interval between the accident or medical treatment and the survey did not have significant effects on the mental health status or clustering results as a standalone confounder.

Still, as you suggest, we can not rule out that the survey results could have been influenced by events after the accident of interest. 
However, in case of the instruments used for quantification of PTSD symptoms, post-traumatic growth, and flashback frequency, our survey indicated that the questions refer to the mountain sport accident. 
For this reason, we assume that the PTS and PTG mental health clusters hallmarked by, respectively, PTSD symptoms and post-traumatic growth represent a long term reaction to the accident. 
We included this information in __Methods__ and __Supplementary Methods__ of the revised manuscript, and we point it also out in __Results/characteristic of the study cohort__ and describe in __Discussion__. 
The possibility that the investigated mental health features are not related to the accident, is stated in the limitation paragraph of __Discussion__.

```{r fig-time, fig.width = figur::convert(rev_figures$time, to = 'in')$w, fig.height = figur::convert(rev_figures$time, to = 'in')$h, fig.cap = 'Effects of ward admission-survey time and of accident year on the major psychometric readouts and the mental cluster assignment.'}

rev_figures$time$plot

```

__Review Figure R\@ref(fig:fig-time). Effects of ward admission-survey time and of accident year on the major psychometric readouts and the mental cluster assignment.__

### 2024-02-08 paper/markdown/manuscript.Rmd
P2P reply, supplementary material
Changes in supplementary figures and tables required by the reviewer's comments

Individuals treated following a mountain sport accident at the Department for Orthopedics and Traumatology at the University Clinic of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 were screened for participation. 
Individuals fulfilling the inclusion criteria: 
hospital admission at least 6 months prior to the start of the study, 
age $\geq$ 18 years, residence in a German-speaking country, and 
proficiency in German (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey via conventional mail. 
Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey. 
Surveys of `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (overall response rate: `r signif(nrow(ptsd$dataset)/nrow(ptsd$cleared), 2) * 100`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).

@@

The psychometric battery consisted of German versions of assessment tools for symptoms of 
PTSD (PCL-5: PTSD checklist for DSM-5) [@Bovin2016], 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ-9: 9-item patient health questionnaire for depressive symptoms) [@Grafe2004], 
panic (PHQ-panic module) [@Grafe2004], 
somatization (PHQ-15: 15-item patient health questionnaire of common somatic symptoms as a substitute for somatization) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
and post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996]. 
The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).

@@

Separate scores were calculated for domains B, C, D and E of the PCL-5 tool along with the total score being the sum of all items. 
Each PCL-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
Significant PTSD symptoms were assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].

Traumatic events prior to the accident were assessed with the DIA-X tool [@Maercker2005]. 
Flashback frequency was surveyed as none, > 1/year and > 1/month. 
Smoking was surveyed as a single yes/no question. 
Problematic alcohol use was investigated with the CAGE tool [@OBrien2008]. 
Data on the type of the mountain sport activity during the accident, injury diagnosis, severity (AIS: abbreviated injury scale) [@Gennarelli2006] and body location, hospitalization, surgery and number of surgical ICD-10 diagnoses were extracted from electronic patient records.

@@

Surveys with complete psychometric data from `r nrow(ptsd$dataset)` individuals were analyzed (__Figure \@ref(fig:fig-consort)__). 
The median time between the trauma center admission after the accident and the survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and fewer surgery diagnoses (__Supplementary Table S4__). 
The survey participants excluded from the analysis due to missingness of psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__). 
Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).

>>>>

Individuals treated following a mountain sport accident at the Department for Orthopedics and Traumatology at the University Clinic of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 were screened for participation. 
Individuals fulfilling the inclusion criteria: 
hospital admission at least 6 months prior to the start of the study, 
age $\geq$ 18 years, 
residence in a <span custom-style = 'revision'>German-speaking regions (Germany, Austria, Switzerland and South Tyrol/Alto Adige in Italy)</span>, and 
proficiency in German (n = `r nrow(ptsd$cleared)`) 
were invited to participate in the online study survey via conventional mail. 
Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey. 
Surveys of `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (overall response rate: `r signif(nrow(ptsd$dataset)/nrow(ptsd$cleared), 2) * 100`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).

@@

The psychometric battery consisted of German versions of assessment tools for symptoms of 
PTSD (PCL-5: PTSD checklist for DSM-5) [@Bovin2016], 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ-9, <span custom-style = 'revision'>PHQ-2 and PHQ-8: 9-, 8- and 2-item patient health questionnaires</span> for depressive symptoms) [@Grafe2004], 
panic (PHQ-panic module) [@Grafe2004], 
somatization (PHQ-15: 15-item patient health questionnaire of common somatic symptoms as a substitute for somatization) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
and post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996]. 
<span custom-style = 'revision'>Items of the PTGI and PTSD instruments were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').</span>
The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).

@@

Separate scores were calculated for domains B, C, D and E of the PCL-5 tool along with the total score being the sum of all items. 
Each PCL-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
Significant PTSD symptoms were assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains. 
<span custom-style = 'revision'>Manifest PTSD was considered for participants positive for all four PCL-5 domains [@Bovin2016].</span>

Traumatic events prior to the accident were assessed with the DIA-X tool [@Maercker2005]. 
<span custom-style = 'revision'>Flashbacks of the accident of interest during mountain sport activity were surveyed as</span> none, > 1/year and > 1/month. 
Smoking was surveyed as a single yes/no question. 
Problematic alcohol use was investigated with the CAGE tool [@OBrien2008]. 
Data on the type of the mountain sport activity during the accident, injury diagnosis, severity (AIS: abbreviated injury scale) [@Gennarelli2006] and body location, hospitalization, surgery and number of surgical ICD-10 diagnoses were extracted from electronic patient records.

@@

Surveys with complete psychometric data from `r nrow(ptsd$dataset)` individuals were analyzed (__Figure \@ref(fig:fig-consort)__). 
The median time between the trauma center admission after the accident and the survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and fewer surgery diagnoses (__Supplementary Table S4__). 
The survey participants excluded from the analysis due to incomplete psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__). 
Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).
